Competing Bandits: The Perils of Exploration under Competition

### 2018-09-02 paper/content/inverted_u.tex

\documentclass[../competing_bandits.tex]{subfiles}
\begin{document}
\section{Competition vs Better Algorithms}\label{section:5}
\swcomment{how about just name the section ``inverted U''?}  \swedit{Now
  we demonstrate the inverted-U relationship between the incentivized
  algorithms and the competition levels. In particular, we investigate
  the incentivized algorithms under three different regimes: permanent
  monopoly, temporary monopoly, and permanent duopoly. \gaedit{We break indifference between exploration algorithms and non-exploration algorithms towards non-exploration algorithms.} First, the story for permanent monopoly is relatively straightforward.}

\swedit{\xhdr{Permanent Monopoly.} Since there is only a single firm
  in the market for the entire period, the firm can take the entire
  market regardless what algorithm it deploys. Assuming the deployment
  of any exploration algorithms will incur certain cost, the firm
  would choose to deploy $\DG$.}

@@

\swdelete{We interpret the move from permanent monopoly to temporary monopoly and temporary monopoly to permanent duopoly as increases in competition. We believe this is reasonable and consistent with the empirical results which define increases in competition as increases in a measure of market power (1 - the Lerner Index). In our model we have no prices and no costs and so we interpret increases in ``market power" as decreases in the number of rounds where a firm has to compete for agents.

% couldn't get section references to work so hardcoding for now
We utilize the same instances and realizations as in the section 3  and simulate the model described in section 2. We initially take the strategies of the firms as exogenous (chosen from $\mathcal{A}$) and simulate the model given these strategies in order to determine the expected payoffs associated with each pair of algorithms. Unless otherwise noted, all the results are reported at $t = 2000$.

\xhdr{Permanent Monopoly} Under permanent monopoly only a single firm exists in the market for every period. In our model the firm only gets utility from a larger market share so it is indifferent between each algorithm in $\mathcal{A}$ since, regardless of the algorithm it deploys, it will get the entire market. If we suppose that deploying the ``better" algorithms has even an $\epsilon > 0$ cost or, equivalently, that firms break indifference towards algorithms that are easier to deploy, then the firm would choose to deploy $\DG$.
}

@@

\OMIT{
\begin{definition}
\textit{Effective End of Game (EEOG)} - the last round, $t$, in a simulation where the agent alive at $t-1$ and the agent alive at $t$ choose different firms.
\end{definition}
}
Since both firms enter at the same time the game is symmetric and we only need to compute the payoffs of three pairs of strategies. We do not report them in the figures, but when both firms play the same algorithm the expected market share is 50/50.

\begin{finding}
\textit{
\begin{enumerate}
\item For low warm starts, under Heavy Tail and Uniform \\ $\DG > \TS$ in competition (see Table \ref{sim_ht}),
\item Under Needle In Haystack, \\ $\TS > \DG$ (see Table \ref{sim_nih}). 
\item For sufficiently large warm starts, $\TS$ is dominant across all instances.
\item Even for a large warm start, $\DG > \DEG$ under Heavy Tail
\end{enumerate}
}
\end{finding}

@@

Combining this finding with the fact that the EEOG values are relatively low across all warm starts \footnote{Interestingly, the EEOG also appears to be skewed to the right}, these observations seem to imply that understanding the performance of different algorithms in competition it is important to look at their performance for relatively small samples instead of asymptotically.

@@

\begin{table}[ht]
\centering
\caption{Duopoly Experiment Needle In Haystack}
\begin{tabular}{rlll}
  \hline
 & $T_0$ = 20 & $T_0$ = 250 & $T_0$ = 500 \\
  \hline
TS vs DG & \makecell{\textbf{0.64} $\pm$0.03\\ eeog \\ avg: 200\\ med: 27} & \makecell{\textbf{0.6} $\pm$0.03\\ eeog \\ avg: 370\\ med: 0} & \makecell{\textbf{0.64} $\pm$0.03\\ eeog \\ avg: 580\\ med: 121.5} \\
  TS vs DEG & \makecell{\textbf{0.57} $\pm$0.03\\ eeog \\ avg: 150\\ med: 14} & \makecell{\textbf{0.52} $\pm$0.03\\ eeog \\ avg: 460\\ med: 78.5} & \makecell{\textbf{0.56} $\pm$0.02\\ eeog \\ avg: 740\\ med: 627.5} \\
  DG vs DEG & \makecell{\textbf{0.46} $\pm$0.03\\ eeog \\ avg: 340\\ med: 128.5} & \makecell{\textbf{0.42} $\pm$0.02\\ eeog \\ avg: 650\\ med: 408} & \makecell{\textbf{0.42} $\pm$0.02\\ eeog \\ avg: 690\\ med: 466.5} \\
   \hline
\end{tabular}
\label{sim_nih}
\caption*{\tiny{The first line in each cell contains the average market share received by the firm playing Alg 1 (and the market share of Alg 2 is 1 - Alg 1 Market Share) as well as a 95 \% confidence band. For example, the cell in the top left indicates that TS gets on average 64\% of the market when played against DG. The next line contain the average and median effective end of game for this set of simulations.}}
\end{table}
% latex table generated in R 3.4.0 by xtable 1.8-2 package
% Thu Aug 16 13:13:00 2018

\xhdr{Temporary Monopoly} We now consider asymmetries in the timing of entry so that one firm enters the market before the other and serves as a monopolist in the periods until the other firm enters. In terms of our model, this corresponds to varying the value of $X$. For this section we report results fixing $T_0 = 20$ and focus on the family of instances where we observed that, for this warm start, $\DG > \TS$ \footnote{However, we report the results for the Needle In Haystack family in the supplementary material. The results are unsurprising, where $\TS$ is the dominant strategy for both the entrant and incumbent.}. It seems clear that the temporary monopoly will get a larger market share regardless of what algorithm it plays, but is the incumbent incentivized to commit to $\TS$?

@@

In table \ref{ht_incum} we can see that, for $X = 200$, $\TS$ is a dominant strategy for the incumbent and that $\DG$ is the dominant strategy for the entrant on the Heavy Tail family \footnote{The result for the Uniform is the same, though it takes a higher $X$ to induce the incumbent to play $\TS$.}. In the supplementary material we report the same experiment for different values of $X$ and show that our main result that, for sufficiently large $X$, $\TS$ is the dominant strategy for the incumbent is robust across the family of instances we consider.

>>>>

\documentclass[../competing_bandits.tex]{subfiles}
\begin{document}
\section{Competition vs Better Algorithms: Inverted-U}\label{section:5}
 Now we demonstrate the inverted-U relationship between the incentivized
  algorithms and the competition levels. In particular, we investigate
  the incentivized algorithms under three different regimes: permanent
  monopoly, temporary monopoly, and permanent duopoly. \gaedit{We break indifference between exploration algorithms and non-exploration algorithms towards non-exploration algorithms due to deployment costs.} \gadelete{First, the story for permanent monopoly is relatively straightforward.}
 We summarize the relationship between the different competition levels and incentivized algorithms as follows:

\begin{conjecture}\label{conj:inverted-u}
\textit{For sufficiently low warm start,
 the optimal strategies in the competition game are:
\begin{center}
\textbf{Permanent Monopoly} - $\DG$ \\
\textbf{Temporary Monopoly} - $\TS$ \\
\textbf{Permanent Duopoly} - $\DG$ \\
\end{center}
The conditions on $\TS$ being dominant under the temporary monopoly are that the incumbent is a temporary monopoly for sufficiently many periods.
}
\end{conjecture}

\xhdr{Permanent Monopoly.} Since there is only a single firm
  in the market for the entire period, the firm can take the entire
  market regardless what algorithm it deploys. \gadelete{Assuming the deployment of any exploration algorithms will incur certain cost} \gaedit{Since we break indifferences by supposing the firm prefers easier to deploy algorithms}, the firm would choose to deploy $\DG$.

@@

@@

Since both firms enter at the same time the game is symmetric and the competition game is zero-sum so we only need to compute the expected payoffs of three pairs of strategies. We do not report them in the figures, but when both firms play the same algorithm the expected market share is 50/50. Since our results are from simulations, we interpret algorithm $A$ being better than algorithm $B$ if the resulting market share of $A$ is statistically greater than $0.5$. Otherwise, the firm is indifferent between $A$ and $B$ with ties broken towards algorithms that are easier to deploy.

Tables \ref{sim_ht} and \ref{sim_nih} show the results of the simulations. For the most part Conjecture~\ref{conj:inverted-u} is validated where, for the Heavy Tail and Uniform instances, we see that $\DG$ does better than $\TS$ and is a weakly dominant strategy in the game. However, there are two results that are at odds with the conjecture:

\begin{puzzle}\label{puz:dg_ht}
Even for large warm start, $\DG$ is better than $\DEG$ under Heavy Tail
\end{puzzle}

\begin{puzzle}\label{puz:nih}
Under Needle In Haystack, $\TS$ is weakly dominant across all warm starts
\end{puzzle}

\OMIT{
\begin{finding}
\textit{
\begin{enumerate}
\item For low warm starts, under Heavy Tail and Uniform \\ $\DG$ is a weakly dominant strategy in competition (see Table \ref{sim_ht}),
\item  (see Table \ref{sim_nih}). 
\item For sufficiently large warm starts, $\TS$ is dominant across all instances.
\item Even for a large warm start, $\DG$ leads to a higher expected market share than $\DEG$ under Heavy Tail
\end{enumerate}
}
\end{finding}
}

@@

We defer the explanation for these puzzles until the next section. Looking at the EEOG values we note that, even though the game continues until $t=2000$, the game effectively ends much earlier than that. This seems to imply that when attempting to understand the performance of different algorithms in competition it is important to look at their performance for relatively small samples instead of asymptotic performance. This is due to the fact that the losing firm may enter into a ``death spiral" where the winner keeps learning and improving from its customers thus increasing its reputation while the losing firm gets starved of customers and progressively falls farther and farther behind its competitors.

@@

\begin{table}[ht]
\centering
\caption{Duopoly Experiment Needle In Haystack}
\begin{tabular}{rlll}
  \hline
 & $T_0$ = 20 & $T_0$ = 250 & $T_0$ = 500 \\
  \hline
TS vs DG & \makecell{\textbf{0.64} $\pm$0.03\\ eeog \\ avg: 200\\ med: 27} & \makecell{\textbf{0.6} $\pm$0.03\\ eeog \\ avg: 370\\ med: 0} & \makecell{\textbf{0.64} $\pm$0.03\\ eeog \\ avg: 580\\ med: 121.5} \\
  TS vs DEG & \makecell{\textbf{0.57} $\pm$0.03\\ eeog \\ avg: 150\\ med: 14} & \makecell{\textbf{0.52} $\pm$0.03\\ eeog \\ avg: 460\\ med: 78.5} & \makecell{\textbf{0.56} $\pm$0.02\\ eeog \\ avg: 740\\ med: 627.5} \\
  DG vs DEG & \makecell{\textbf{0.46} $\pm$0.03\\ eeog \\ avg: 340\\ med: 128.5} & \makecell{\textbf{0.42} $\pm$0.02\\ eeog \\ avg: 650\\ med: 408} & \makecell{\textbf{0.42} $\pm$0.02\\ eeog \\ avg: 690\\ med: 466.5} \\
   \hline
\end{tabular}
\label{sim_nih}
\caption*{\tiny{The first line in each cell contains the average market share received by the firm playing Alg 1 (and the market share of Alg 2 is 1 - Alg 1 Market Share) as well as a 95\% confidence band. For example, the cell in the top left indicates that TS gets on average 64\% of the market when played against DG. The next line contain the average and median effective end of game for this set of simulations.}}
\end{table}
% latex table generated in R 3.4.0 by xtable 1.8-2 package
% Thu Aug 16 13:13:00 2018

\xhdr{Temporary Monopoly} We now consider asymmetries in the timing of entry so that one firm enters the market before the other and serves as a monopolist in the periods until the other firm enters. In terms of our model, this corresponds to varying the value of $X$. For this section we report results fixing $T_0 = 20$ and focus on the family of instances where we observed that, for this warm start, $\DG$ was the optimal strategy \footnote{However, we report the results for the Needle In Haystack instances in the supplementary material. The results are unsurprising, where $\TS$ is the dominant strategy for both the entrant and incumbent.}. It seems clear that the temporary monopoly will get a larger market share regardless of what algorithm it plays, but is the incumbent incentivized to commit to $\TS$?

@@

In table \ref{ht_incum} we can see that, for $X = 200$, $\TS$ is a dominant strategy for the incumbent and that $\DG$ is the dominant strategy for the entrant on the Heavy Tail instances \footnote{The result for the Uniform instances is the same, though it takes a higher $X$ to induce the incumbent to play $\TS$.}. In the supplementary material we report the same experiment for different values of $X$ and show that our main result that, for sufficiently large $X$, $\TS$ is the dominant strategy for the incumbent is robust across the instances we consider.

### 2018-09-02 paper/content/revisited.tex

The results from the previous section present a few puzzles:
\begin{enumerate}
\item $\DG > \DEG$ on Heavy Tail even for large warm starts does not seem consistent with the ordering of the means.
\item Why does $\TS$ beat $\DG$ for Needle In Haystack for low warm starts whereas on Uniform and Heavy Tail it does not?
\end{enumerate}

We first attempt to explain the first puzzle. We revisit Conjecture~\ref{conj:mean-trajectories} and find a more nuanced picture.

@@

Figure \ref{relative_rep_plots} shows plots of the relative reputation proportion for $\TS$ vs $\DG$ on the Uniform and Needle In Haystack instances. For the Uniform instance we see that, in the early rounds, $\DG > \TS$ for the majority of the simulations but that, eventually, $\TS > \DG$. The intuition behind this is that, especially since the firms start with no substantive initial information, $\TS$ does purposeful exploration in the early rounds in order to acquire information. However, eventually the information acquired from purposeful exploration in the early rounds allows $\TS$ to make better decisions and achieve a higher reputation, especially when the instance is ``hard enough" so that $\DG$ cannot trivially find the best arm. The early exploration leads to what we define as a \textit{relative reputation cost} and the eventual gain in reputation leads to what we define as a \textit{relative reputation benefit}.

\begin{definition}
\textit{Relative reputation cost and benefit} - the relative reputation loss an algorithm incurs from purposeful exploration compared to the greedy alternative. Here we treat it as an empirical definition based on the relative reputation proportion where the ``cost" regime is when the proportion $< 0.5$ and the ``benefit" regime is when the proportion is $> 0.5$ for the better algorithm. We call the priors where there is an early costly period followed by a later benefit period relative reputation costly instances.
\end{definition}
Is exploration always costly? Figure \ref{relative_rep_plots} also shows that for the Needle In Haystack instances, $\TS$ always does relatively better than $\DG$. There are two contributing factors to this. First, $\TS$ identifies the best arm faster in the Needle In Haystack prior than the Uniform prior so that there is a shorter time horizon where $\TS$ needs to engage in purposeful exploration. Second, in the Needle In Haystack prior there are no ``bad" arms as there may be in the Uniform prior since by construction all the arms except one in Needle In Haystack are the same. Thus, when $\TS$ pulls a sub-optimal arm relative to its current information, the expected reward is the same as the greedy option that has not identified the best arm. However, with the Uniform prior, it is possible that the sub-optimal arm that is pulled has substantially lower expected reward relative to the greedy option. Thus, only the Uniform and Heavy Tail instances are relative reputation costly priors.

This observation allows us to explain the second puzzle from the competition game with a more nuanced view of the inverted-U.

\begin{finding}
\textit{For sufficiently low warm start on relative reputation costly instances the optimal strategies in the competition game are:
\begin{center}
\textbf{Permanent Monopoly} - $\DG$ \\
\textbf{Temporary Monopoly} - $\TS$ \\
\textbf{Permanent Duopoly} - $\DG$ \\
\end{center}
The conditions on $\TS$ being dominant under the temporary monopoly are that the incumbent is a temporary monopoly for sufficiently many periods.
}
\end{finding}

>>>>

Conjecture~\ref{conj:inverted-u} for the most part is validated by our simulation results, but Puzzles~\ref{puz:dg_ht} and \ref{puz:nih} seem to be problematic for it.

We first attempt to explain Puzzle~\ref{puz:dg_ht} by revisiting Conjecture~\ref{conj:mean-trajectories}.

@@

Figure \ref{relative_rep_plots} shows plots of the relative reputation proportion for $\TS$ vs $\DG$ on the Uniform and Needle In Haystack instances. For the Uniform instance we see that, in the early rounds, the relative reputation proportion for $\TS$  vs. $\DG$ is less than 0.5 meaning that $\DG$ has a higher reputation in a majority of the simulations  but that, eventually, the proportion $\TS$ does better than $\DG$ in a majority of the simulations. The intuition behind this is that, especially since the firms start with no substantive initial information, $\TS$ does purposeful exploration in the early rounds in order to acquire information. However, eventually the information acquired from purposeful exploration in the early rounds allows $\TS$ to make better decisions and achieve a higher reputation, especially when the instance is ``hard enough" so that $\DG$ cannot trivially find the best arm. The early exploration leads to what we define as a \textit{relative reputation cost} and the eventual gain in reputation leads to what we define as a \textit{relative reputation benefit}.

\begin{definition}
\textit{Relative reputation cost and benefit} - the relative reputation loss an algorithm incurs from purposeful exploration compared to the greedy alternative. Here we treat it as an empirical definition based on the relative reputation proportion where the ``cost" regime is when the proportion $< 0.5$ and the ``benefit" regime is when the proportion is $> 0.5$ for the better algorithm. We call the priors where there is an early costly period followed by a later benefit period relative reputation costly instances.
\end{definition}

Is exploration always costly? Figure \ref{relative_rep_plots} also shows that for the Needle In Haystack instances, $\TS$ always does relatively better than $\DG$. There are two contributing factors to this. First, $\TS$ identifies the best arm faster in the Needle In Haystack prior than the Uniform prior so that there is a shorter time horizon where $\TS$ needs to engage in purposeful exploration. Second, in the Needle In Haystack prior there are no ``bad" arms as there may be in the Uniform prior since by construction all the arms except one in Needle In Haystack are the same. Thus, when $\TS$ pulls a sub-optimal arm relative to its current information, the expected reward is the same as the greedy option that has not identified the best arm. However, with the Uniform prior, it is possible that the sub-optimal arm that is pulled has substantially lower expected reward relative to the greedy option. Thus, only the Uniform and Heavy Tail instances are relative reputation costly priors.

This observation allows us to explain the Puzzle~\ref{puz:nih} from the competition game and establish a more nuanced view of the inverted-U.

\begin{finding}
\textit{For sufficiently low warm start \emph{under the instances that are relative reputation costly}, the optimal strategies in the competition game are:
\begin{center}
\textbf{Permanent Monopoly} - $DG$ \\
\textbf{Temporary Monopoly} - $\TS$ \\
\textbf{Permanent Duopoly} - $DG$ \\
\end{center}
The conditions on $\TS$ being dominant under the temporary monopoly are that the incumbent is a temporary monopoly for sufficiently many periods.
}
\end{finding}
